---
title: |
  | \vspace{5cm} Lab 1: Question 1: Are Democrats voters older or younger than Republican voters in 2020?
author: "Elaine Chang, Dom Dillingham, Jesse Miller, Michael Wang"
output: pdf_document
---

\newpage 


```{r load packages, echo=TRUE, include=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2) 
library(tidyverse)
library(magrittr)
library(knitr)
```


## 1. Importance and Context
Politics often use basic demographics as a lens to better understand and plan around voters. Political operatives - e.g., current / aspiring politicians, political consultants, third-party watchdogs to name a few - may use the variable "age" as a specific demographic field to better understand their voters and highlight differences between the two major parties. This can impact strategic functions such as policy formation or tactical initiatives such as marketing outreach so as to better engage voters. For example, Democrats have taken up \href {https://www.forbes.com/sites/adamminsky/2020/09/21/how-cancel-student-debt-went-from-a-fringe-idea-to-mainstream/?sh=448fc4045489}{\bf canceling student debt} as part of their policy mantle with institutional support very recently, though historically it sat as a fringe idea, popular amongst Millennials. Thus understanding if there is an age difference, and further  what the age difference is, between Democratic and Republican voters, can help political operatives better navigate their party towards a path to power.

## 2. Description of Data
The data used in this analysis is from the preliminary 2020 American National Election Studies (ANES) Time Series Study, released February 11, 2021. The ANES 2020 Time Series Study reflects a total of 8,280 pre-election interviews conducted by one of the three mode groups - by web, video or telephone. Readers are encouraged to refer to the \href{https://electionstudies.org/data-center/2020-time-series-study/}{\bf ANES full codebook} for additional queries on the original data that is available online.

In operationalizating our research question ("Are Democrats voters older or younger than Republican voters in 2020?"), we sought to clearly define the variables we were most interested in investigating. While the ANES study encompasses a dataset of over 700 different variables, we identified three variables in ANES that is most pertinent to investigating our research question.

a) Age. The survey asked respondents for their age, which will be used as the primary response variable. We identified variable "V201507x" as the most suitable variable. Any responses of "-9" which maps to "Refused" were filtered out of the dataset due to the core consideration factor that age in the research question. Age was reported in the survey as their true value from ages 18, the youngest eligible age to vote, to age 79. Respondents over the age of 80 were grouped together and reported as one value. We focused this analysis on those aged 18 to 79. While excluding the 80+ age group may introduce some bias in the analysis, we expect the contribution of the 80+ group to the overall age distribution to be relatively small because the total number of Democrats and Republicans respondents over 80 years old are similar in total count (178 and 195, respectively).

```{r examine age variable,echo=FALSE, warning=FALSE,message=FALSE}
#nrow(dat[dat$Voting_Party == 'Democrat' & dat&Age == 80, ])
#nrow(dat[dat$Voting_Party == 'Republican' & dat&Age == 80, ])
```

b) Party Affiliation. In defining "Democrat" and "Republican" groups, we relied on the respondent's preference for presidential candidate (captured in V201075x). There were several contending interpretations to define party affiliation such as party registration (V201018 "Party of registration"), self-identification (V201228  "Does R think of self as Democrat, Republican, or Independent?"), or preference for a selected office (e.g., V201070 "For whom does R intend to vote for governor?").  However our qualitative research found that party registration may be an unreliable indicator. For example, an \href{ https://www.dcreport.org/2020/12/19/mitch-mcconnells-re-election-the-numbers-dont-add-up/}{\bf analysis by DC Report} of Mitch McConnell's recent 2020 re-election victory discusses that conventional political wisdom warns that "analysts shouldn't correlate party registration with voting patterns." Similarly, we ruled out the party self-identification variables as being too subjective on voter interpretation, without the specificity of action weighed with it. Finally, we chose a variable that tied action and preference to the tangible preference of the presidential office. Down ballot (e.g., Congressional) and local offices could introduce complications on state or city context considerations into the analysis. We are focusing on the presidential candidate position specifically as that position is chosen as the national leader by their respective party. Additionally, the presidential candidate for each party increasingly define boundaries of the party identity such as Donald Trump and the GOP. 

c) Voters. For the purposes of this survey, our team wanted to pointedly analyze the fullest scope of voters with an emphasis on respondents who will cast a ballot. However because this data is pre-election, it is unlikely to be accurate in its capture of respondents as voters at the time of data collection. When the full data set for 2020 is released by ANES in a future date, we could re-analyze the data. In the meantime, we defined voters by capturing respondents who voted, intended to vote or preferred a presidential candidate (V201075x) with an additional filter of those who also intended to register to vote (V201019). Together, this implied enough knowledge and action on behalf of the respondent to create our definition of "voter."

We plot the ages of voters in the Democrat and Republican groups in Figure 1 to generate basic observations of the dataset. We observe a relatively greater population of younger voters in the Democrat group, although tests need to be performed to determine statistical significance. We also note that the data in both groups do not appear skewed, and when combined with the large data count, assures general normality of distributions.

```{r load and clean data, echo=FALSE, include = TRUE, message=FALSE, fig.height=4, fig.width=10, fig.cap="\\label{fig:figs}Age Distributions of Democratic and Republican Voters"}
rawDat <- haven::read_dta("anes_timeseries_2020_stata_20210211/anes_timeseries_2020_stata_20210211.dta")

## Select fields
dat <- rawDat %>% select(V201507x,V201075x,V201019) %>%
  rename(Age = V201507x,
         Voting_Party = V201075x,
         register = V201019) %>%

## Group party affiliation and voting intent/preference and action
mutate(Voting_Party = case_when(Voting_Party %in% c(10, 20, 30) ~ 'Democrat',
                           Voting_Party %in% c(11, 21, 31) ~ 'Republican',
                           TRUE ~ 'Other'))

# V201507x PRE: SUMMARY: RESPONDENT AGE
# Values
# 18-79 == Actual value
# 80 == Age 80 or older
# -9 == Refused

# V201075x PRE: SUMMARY: PARTY OF PRE-ELECTION
# PRESIDENTIAL VOTE/INTENT/PREFERENCE
# Values
# 10 == Democratic candidate selected (vote)
# 11 == Republican candidate selected (vote)
# 12 == Other candidate selected (vote)
# 20 == Democratic candidate selected (intent to vote)
# 21 == Republican candidate selected (intent to vote)
# 22 == Other candidate selected (intent to vote)
# 30 == Democratic candidate selected (preference)
# 31 == Republican candidate selected (preference)
# 32 == Other candidate selected (preference)
# -1 == Inapplicable

dat %<>% filter(!(Voting_Party %in% c(-1)),
                !(Age %in% c(-9)),
                  register %in% c(1, -1))

# hist_D<-hist(dat[dat$Voting_Party == 'Democrat', ]$Age,main="Democratic Voters by Age",xlab="Age",col="blue")
# hist_R<-hist(dat[dat$Voting_Party == 'Republican', ]$Age,main="Republican Voters by Age",xlab="Age",col="red")

par(mfrow=c(1, 2))
hist(dat[dat$Voting_Party == 'Democrat', ]$Age,main="Democratic Voters by Age",xlab="Age",col="blue")
hist(dat[dat$Voting_Party == 'Republican', ]$Age,main="Republican Voters by Age",xlab="Age",col="red")

```


## 3. Most appropriate test 
We are interested in whether the average age of a Democratic voter differs from the average age of a Republican voter. As such, we determined that the most appropriate statistical test is the Welch two-sample two-sided t-test. This will be an unpaired t-test as the variables are independent with no known systemic dependencies. It satisfactorily fulfills the following core requirements of the Welch two-sample two-sided t-test: 

a) Metric data. The variable we are assessing, Age, is typically measured on a metric scale. However, the variable measuring Age (V201507x) captures Age on a metric scale from 18-79, which is on a metric scale but the survey groups all 80+ ages in one large bucket which is not metric. As described in the above section 3, we censored this upper side of the Age scale (80+) in our analysis.

b) Approximately normal distribution. Figure 1 shows voter age distributions across the two parties. While the histograms are not a definitively normal distribution, they also do not show extreme skewness. If we had a small sample size (i.e., less than 30) we could consider the nonparametric Wilcoxon Rank-Sum test. However because, the sample size is large enough such that the Central Limit Theorem can be applied to assert a normal distribution, we proceeded with the Welch t-test. 

c) Data drawn IID. It is unlikely that the sample of survey respondents drawn and reported on fully satisfy the requirement of IID. Per the ANES userguide / codebook, weights are to be used to mitigate sampling bias identified by the study's researchers. The large sample size drawn across the 50 states provide some assurance in some partial fulfillment of IID. Additionally, per instructions for this research assignment, weights will be ignored. Therefore, we will proceed with our analysis assuming the data fulfills the IID requirement.

In addition, it is worth noting that we explicitly chose to continue with Welch's t-test as opposed to the Student's t-test. Similar to many different statistical programs, R’s default t-test is set to the Welch t-test. In designing our test, there was no obvious indication to alter the default Welch t-test in R the Student's t-test. We would rather withhold applying  assumptions on the sample variances. \href{http://daniellakens.blogspot.com/2015/01/always-use-welchs-t-test-instead-of.html}{Cursory reports by others} interested in the differences between these two t-tests show that the Welch t-Test performs better than the Student’s t-Test when sample sizes and variances are unequal between groups but gives identical results when variances are equal. Thus there does not appear to be an advantage in making upfront assumptions regarding variances and move away from the Welch t-test.

Our hypotheses for this test will be: 

* H0: The average age of Democratic voters and the average age of Republican voters are the same 

* Ha: The average age of Democratic voters and the average age of Republican voters are not the same

\


```{r test, echo=TRUE, warning=TRUE, message=TRUE}
d = dat[dat$Voting_Party == 'Democrat' ,]$Age
r = dat[dat$Voting_Party == 'Republican' ,]$Age
t.test(d,r)

```
## 4. Test, results and interpretation
The results of our Welch two-sample t-test show that the p-value is 2.036e-13. Comparing this p-value to the industry-accepted alpha level of 0.05 on a two-sided test, this test suggests that there is a "very highly significant" statistical finding to reject the null hypothesis that Democrats and Republicans have the same average age. Statistically, out of all confidence intervals constructed according to our procedure outlined, we are 95% confident that Democrats are 2.2 to 3.8 years younger than their Republican counterparts.

While statistically there is strong evidence to suggest a difference in average age between the two party voters, there seems to be limited practical application of this procedure and result. The null hypothesis we rejected was that Democrat voters and Republican voters do not have the same average age. We can infer based on this that Democrat voters are indeed younger than Republican voters. However the test also shows that the actual difference in age is a relatively narrow margin of approximately three years on average.  Thus on a practical level, the application of this knowledge is limited. Policy direction, candidate messaging, marketing outreach and the like will not be swayed using a strictly age-related lens if the basis of that difference is three years between the two political parties.

## 5. Limitations
There were several limitations to the data and analysis presented that should be noted. 

* It should be emphasized that this dataset is both a recent and preliminary release of the 2020 ANES Time Series Study. As the data is updated, our results may be updated. 

* As mentioned earlier, weights were encouraged by the ANES user guide and those were not applied, per assignment instructions and time limitations. 

* There are two macro events to highlight in 2020 that could cause irregularities in the data. Firstly, COVID-19 impacted this study in multiple arenas from data collection to survey questions to address the challenges brought by the pandemic. Secondly, the 2020 election was highly unusual due to then-President Trump's false accusations in the untrustworthiness of the vote-by-mail process. Trump's repeated claims may have caused respondents to opt-out of voting altogether and this would impact our "voters" group in the analysis. 