---
title: 'Lab 1: Question 2'
author: "Elaine Chang, Dom Dillingham, Jesse Miller, Michael Wang"
output: pdf_document
---

```{r load packages, echo=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(tidyverse) # if you want more, but still core, toolkit
library(magrittr)
library(reshape2)
library(knitr)

# Load dplyr last for any package issues
library(dplyr)
```

```{r load and clean data, echo=FALSE, warning=TRUE, message=FALSE}
rawDat <- haven::read_dta('anes_timeseries_2020_stata_20210211/anes_timeseries_2020_stata_20210211.dta')

## Select fields & create understandable features
dat <- rawDat %>% select(V201153, V201151, V201018, V201075x, V201228, V200010a) %>%
  rename(Kamala_Meter = V201153,
         Biden_Meter = V201151,
         Party = V201018,
         Voting_Party = V201075x,
         weight = V200010a) %>%
  mutate(Voting_Party = case_when(Voting_Party %in% c(10, 20, 30) ~ 'Democrat',
                           Voting_Party %in% c(11, 21, 31)~ 'Republican',
                           TRUE ~ 'Other'))

dat %<>% filter(Voting_Party == 'Democrat',
                !(Kamala_Meter %in% c(998, 999, -4, -9)),
                !(Biden_Meter %in% c(998, 999, -4, -9)))
```


## Importance and Context
In this analysis, we will be exploring whether Democratic voters are more enthusiastic about Joe Biden or Kamala Harris. With a contentious Democratic primary, we believe this exploration can help explore whether the selection of Kamala Harris may have made the Democratic ballot more positive to those voting Demomcratic.

## Description of Data
For this analysis, we will operationalize a Democratic voter to be whether the respondent had voted, intention to vote, or preference for a Democratic presidential candidate. This field is a summary field calcuated by the American National Election Studies. Further, we will operationalize a respondent's preference for Joe Biden or Kamala Harris with a respective `Feeling Thermometer` where the respondent ranks their feeling on a scale from 0 to 100.

The operationalization of a Democratic voter is one of interpretation. Instead of an individuals intent or preference to vote, we could also have operationalized with a respondent's party of regristration. While not all voters are registered, we beleive the selected variable is more appropriate.

The use of the `Feeling Thermometer` does appear to be an appropriate variable for this study. This will allow for better distinction between an individual's perception of each candidate.

For this analysis, the total dataset was just filtered down to those voting for a Democratic presidential candidate. Further, any erroneous values for either the Joe Biden or Kamala Harris thermometer are removed.

From Table 1 below, we see that the difference in thermometer between Joe Biden and Kamala Harris is very close, with Joe Biden being higher by just 1.6 points. While there does appear to be preference for Joe Biden, it is not by much.

```{r distribution table, echo=FALSE, warning=TRUE, message=FALSE, results = 'asis'}
tableDat <- dat %>% mutate(diff = Biden_Meter - Kamala_Meter) %>% summarise(`Record Count` = n(),
                              `Average Harris Thermometer` = round(mean(Kamala_Meter), 2),
                              `Average Biden Thermometer` = round(mean(Biden_Meter), 2),
                              `Average Difference` = round(mean(diff), 2))
kable(tableDat, caption = 'Data Distribution by Politician')
```

From the table below, we see that the difference between Joe Biden and Kamala Harris is fairly symmetric with a large point-mass around zero where respondants equally fair both politicians.  

```{r graphic distribution type2, echo=FALSE, warning=TRUE, message=FALSE}
dat %<>% mutate(diff =  Biden_Meter - Kamala_Meter)

ggplot(dat, aes(x=diff)) +
  geom_histogram(color="black", fill="white") +
  geom_vline(aes(xintercept=mean(diff)),
            color="blue", linetype="dashed", size=1) +
  labs(title="Difference Between Joe Biden & Kamala Harris",
       x="Difference in Thermometer",
       y="Number of Responses")

```

## Most appropriate test
The most appropriate test for this analysis a Wilcoxon Signed Rank Test. For this problem, a paired statistical test is prefered as we are not controlling for the influence of other factors that may impact the difference such as age and historical voting patterns, which a paired test will help account for. Within the range of paired test, the Wilcoxon Signed Rank Test pass all the tests assumptions while providing the most power. The assumptions are:

* Metric scale - The thermometer is on a metric scale. 
* Paired dependent samples - Each of the drawings are paired dependent samples as the same individual would be asked both questions.
* Difference between samples is symmetric about mean - From the graph above, it does appear that the data is symmetric about the mean.
* Data is drawn IID - This data does not fully align with the properties of IID. From the properties of the survey, some individuals are more likely to be sampled and weights to account for this sampling bias were provided. For the purposes of this analysis, these weights were ignored.

## Test, results and interpretation
```{r test, echo=FALSE, warning=TRUE, message=FALSE}
wilcox.test(dat$Biden_Meter, dat$Kamala_Meter, paired = TRUE, alternative = "two.sided")
```
The result of the statisical test indicate that, with an alpha-level of 0.05, we reject the null hypothesis that the difference bewteen thermometers of Joe Biden and Kamala Harris are 0.

With the small measured difference between Kamala Harris and Joe Biden, while there is a statisically significant difference between the politicians, the practical significance is small.
