---
title: 'Lab 1: Question 1'
author: "Your Names Here"
output: pdf_document
---

```{r load packages, echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2) 
# library(tidyverse) # if you want more, but still core, toolkit
```

```{r load and clean data, echo=FALSE, warning=TRUE, message=FALSE}

```



```{R}
library(haven)
raw = read_sav("anes_timeseries_2020_spss_20210211/anes_timeseries_2020_spss_20210211.sav")
# raw_dta = read_dta("anes_timeseries_2020_stata_20210211/anes_timeseries_2020_stata_20210211.dta")
```

```{R}
# V201624: Anyone in household tested pos for COVID-19
# 1 == someone in household tested positive, 287
# 2 == no one tested positive, 7895
# -5 == interview breakoff (sufficient partial IW), 38
# -9 == refused, 60


# V201625: Anyone in household been suspected of having COVID-19 based on symptoms
# 1 == suspected of having COVID, 998
# 2 == no one has been suspected of having COVID, 7185
# -5 == interview breakoff (sufficient partial IW), 38
# -9 == refused, 59

# Of 287 people test positive for COVID, 73 reported no one has been suspected of COVID based on symptoms.
sum(raw[raw$V201624 == 2, ]$V201625 == 1) 

# V201145: Approve/Disapprove of governor's handling of COVID-19
# 1 == Approve, 5083 (Magnitudes: 3212 strongly, 1862 not strongly, 1 don't know, 8 refused)
# 2 == Disapprove, 3145 (Magnitudes: 2251 strongly, 893 not strongly, 1 refused)
# -8 == Don't know, 16 
# -9 == Refused, 36

# V201146: How much Approve/Disapprove of governor's handling of COVID-19
# 1 == Strongly, 5463
# 2 == Not strongly, 2755
# -1 == Inapplicable, 52
# -8 == Don't know, 1
# -9 == Refused, 9

```

Justification for using V201624:  
There are 782 households who have had no one test positive, but suspect someone may have COVID based on symptoms.  
We want to know how  having a family member with COVID affects one's approval of the governor's handling of COVID.
The determination of whether a family member has COVID will rely on the ground truth: testing, not suspicion.
It becomes a different question to ask how someone who thinks a family member may have COVID approves of the governor's handling of COVID.


```{R}
# Approach 1: Use 2 sample T-test and binary (metric) approve/disapprove values.
# binary data IS metric data and ordinal

# V201624: Anyone in household tested pos for COVID-19
# 1 == someone in household tested positive, 287
# 2 == no one tested positive, 7895
# -5 == interview breakoff (sufficient partial IW), 38
# -9 == refused, 60

# V201145: Approve/Disapprove of governor's handling of COVID-19
# 1 == Approve, 5083 (Magnitudes: 3212 strongly, 1862 not strongly, 1 don't know, 8 refused)
# 2 == Disapprove, 3145 (Magnitudes: 2251 strongly, 893 not strongly, 1 refused)
# -8 == Don't know, 16 
# -9 == Refused, 36

binary_T = raw[, c('V201624', 'V201145')]
# all(raw$V201145 == binary_T$V201145)

# Remove invalid values
binary_T = binary_T[(binary_T$V201624 == 1 | binary_T$V201624 == 2), ] # remove -5 and -9 values for test pos for COVID
binary_T = binary_T[(binary_T$V201145 == 1 | binary_T$V201145 == 2), ] # remove -8 and -9 values for binary governor approval

# Normalize from 2 -> 0
# 1 = (V201624) test positive        | (V201145) approve governor
# 0 = (V201145) no one test positive | (V201145) disapprove governor
binary_T[binary_T$V201624 == 2, 'V201624'] = 0
binary_T[binary_T$V201145 == 2, 'V201145'] = 0

# Ho: governor approval value if family infected == governor approval value if no one infected 
# Ha: governor approval value if family infected != governor approval value if no one infected 

# mean(binary_T[binary_T$V201624 == 1, ]$V201145) # test positive, 0.5598592
# mean(binary_T[binary_T$V201624 == 0, ]$V201145) # no one test positive, 0.6208302

test_positive = binary_T[binary_T$V201624 == 1, ]$V201145
test_negative = binary_T[binary_T$V201624 == 0, ]$V201145
t.test(test_positive, test_negative)
# Ho is rejected at a 0.05 significance level. Individuals who have someone test positive are more likely to disapprove if governor.

# Reporting means is not an intuitive measure
library(effsize)
cohen.d(test_positive, test_negative) # d estimate: -0.1255487 (negligible)

cor(binary_T$V201145, binary_T$V201624) # -0.0230376

```


```{R}
# Approach 2: Use Wilcoxon Ranked Sum and (ordinal) Likert (strong approval, non-strong approval, non-strong disapproval, strong disapproval) values.

# V201624: Anyone in household tested pos for COVID-19
# 1 == someone in household tested positive, 287
# 2 == no one tested positive, 7895
# -5 == interview breakoff (sufficient partial IW), 38
# -9 == refused, 60

# V201145: Approve/Disapprove of governor's handling of COVID-19
# 1 == Approve, 5083 (Magnitudes: 3212 strongly, 1862 not strongly, 1 don't know, 8 refused)
# 2 == Disapprove, 3145 (Magnitudes: 2251 strongly, 893 not strongly, 1 refused)
# -8 == Don't know, 16 
# -9 == Refused, 36

# V201146: How much Approve/Disapprove of governor's handling of COVID-19
# 1 == Strongly, 5463
# 2 == Not strongly, 2755
# -1 == Inapplicable, 52
# -8 == Don't know, 1
# -9 == Refused, 9

ordinal_WRS = raw[, c('V201624', 'V201145', 'V201146')]

# Remove invalid values
ordinal_WRS = ordinal_WRS[(ordinal_WRS$V201624 == 1 | ordinal_WRS$V201624 == 2), ] # remove -5 and -9 values for test pos for COVID
ordinal_WRS = ordinal_WRS[(ordinal_WRS$V201145 == 1 | ordinal_WRS$V201145 == 2), ] # remove -8 and -9 values for binary governor approval
ordinal_WRS = ordinal_WRS[(ordinal_WRS$V201146 == 1 | ordinal_WRS$V201146 == 2), ] # remove -1, -8 and -9 values for ordinal governor approval

# Normalize from 2 -> 0
# 1 = (V201624) test positive        | (V201145) approve governor    | (V201146) strongly
# 0 = (V201624) no one test positive | (V201145) disapprove governor | (V201146) not strongly
ordinal_WRS[ordinal_WRS$V201624 == 2, 'V201624'] = 0
ordinal_WRS[ordinal_WRS$V201145 == 2, 'V201145'] = 0
ordinal_WRS[ordinal_WRS$V201146 == 2, 'V201146'] = 0

ordinal_WRS$ordinal_approval = -1
ordinal_WRS[(ordinal_WRS$V201145 == 1) & (ordinal_WRS$V201146 == 1), ]$ordinal_approval = 4 # strong approval
ordinal_WRS[(ordinal_WRS$V201145 == 1) & (ordinal_WRS$V201146 == 0), ]$ordinal_approval = 3 # non-strong approval
ordinal_WRS[(ordinal_WRS$V201145 == 0) & (ordinal_WRS$V201146 == 0), ]$ordinal_approval = 2 # non-strong disapproval
ordinal_WRS[(ordinal_WRS$V201145 == 0) & (ordinal_WRS$V201146 == 1), ]$ordinal_approval = 1 # strong disapproval


test_positive = ordinal_WRS[ordinal_WRS$V201624 == 1, ]$ordinal_approval
test_negative = ordinal_WRS[ordinal_WRS$V201624 == 0, ]$ordinal_approval
wilcox.test(test_positive, test_negative)


```






## Importance and Context
<!-- You can (and should delete each of these comments lines in your final report) --> 
<!-- Explain why the reader should care about your research question. -->

Through 2020, the COVID-19 pandemic affected the lives of millions of Americans. At the time of writing, an estimated 28 million cases have been reported in the US, along with 500,000 deaths, \href{https://covid.cdc.gov/covid-data-tracker/#datatracker-home}{\underline(according to the CDC)}. 

2020 was also an election year, during which gubernatorial (state-level) elections were held in 11 states and 2 territories, concurrently with the presidential election. 9 state governors ran for reelection. 

We attempt to understand whether survey respondents who have had someone in their home infected by COVID-19 were more likely to disapprove of the way their governor is handling the pandemic. Having someone in their home infected by COVID-19 is one of the most intimate ways the pandemic may affect a survey respondent. We want to examine how this affect translates into a survey respondents' perception of their governor's handling of the pandemic. This is crucial, as 9 state governors were running for election in 2020. A marked disapproval could result in a changing power dynamic across many states.



## Description of Data
<!-- Explain how your research question is operationalized, including whether the variables you create are appropriate to study the concepts in question. --> 
<!-- What are some basic features of the data distribution? --> 
<!-- What changes do you make to the data and why? --> 

Raw data originates from the pre-election survey that the American National Election Studies (ANES) conducts, the 2020 Time Series Study. Within this study of more than 700 variables, 4 were identified to be relevant to this analysis. No weighting was performed.

There exists 2 variables that relate to whether a respondent may have been affected by COVID-19 in their household. One records "whether anyone in your household tested positive for COVID-19", while the other asks "whether anyone in your household has been suspected of having COVID-19 based on symptoms". Although related, these variables fundamentally differ. In this analysis, we choose to analyze respondents under the former variable. We believe relying on the ground truth - a positive or negative test result - is most appropriate. With the uncertainty that COVID-19 brings, we want to ensure respondents are not relying on gut feel or suspicion that a member of their household may have COVID-19.

There exists 2 variables that attempt to gauge a respondent's perception of their governor's handling of COVID-19. One variable records a binary approval or disapproval response, while the other variable attempts to describe the magnitude of approval or disapproval, by allowing the respondent to select a strength (strongly, not strongly). We attempt to use both in our analysis.

We remove entries in which respondents broke off an interview, refused to answer, or responded with "Don't know" or "inapplicable" to a survey question. These removed entries constituted 1.7% of our raw data. Some respondents may simply not have a well-formed opinion on their governor or may not have clear knowledge of their household's COVID-19 status. We believe the nature of this non-response is random, and not systematic in nature, so we believe removal of these entries will not bias the results of our analysis.

We normalize binary responses to a scale of 0 to 1.  We create a custom field that combines a respondent's binary approval/disapproval of a governor with the respondent's strength in opinion (strongly, not strongly), which serves as an ordinal variable.

We observe 284 respondents with someone in their household test positive for COVID-19. We observe 7854 respondents without a positive COVID-19 in the household.


## Most appropriate test 
<!-- Explain which test is the most appropriate to answer your question and why it is the most appropriate --> 
<!-- List every assumption that your test requires and evaluate it. -->

## Test, results and interpretation
<!-- What are the results of your test? --> 
<!-- What do the results of this test mean? What is the practical significance? --> 
